10 years of school closures and consolidations: the impact on school accessibility

### 2025-01-12 paper/Manuscript.qmd
Proofread revisions

---
title: "10 years of school closures and consolidations: the impact on school accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
format:
  nature-pdf: 
    journal: "default"
    keep-tex: true
editor: source
execute: 
  freeze: auto

author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca"
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing motorized travel and increasing active travel are top priorities for many communities. However, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools. Public school boards have been strained to do more with less, and opting to close under-capacity schools to deliver educational services more cost-effectively. However, once a school is closed, students' proximity to schools decreases, and those who previously lived close enough to walk or cycle can no longer do so. In this paper, the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced a 10% decrease in elementary school locations between 2011 and 2021, serves as a case study. We quanitfiy how these closures and consolidations have reshaped accessibility for the elementary school-aged population and discuss their implications for active and motorized travel.  To measure accessibility, spatial availability, a singly-constrained multimodal measure of spatial accessibility is used. This method's proportional allocation mechanism enables interpretable comparison between accessibility 'before' and 'after' the implementation of school closure/consolidation policy considering changes in potential travel time and population. The analysis is conducted at the parcel level, considering all school seats as eligible and incorporating observed home-to-school travel patterns for motorized and non-motorized modes. We present system-wide impact estimates resulting from these closures. The findings reveal that overall spatial availability in the city declined. While the school consolidation policy may have achieved cost savings in facility operation and maintenance, the burden falls on students, families, and communities through _potentially_ reduced opportunities for physical activity through an active journey to school and increased motorized travel externalities.

@@

```{r, include=FALSE}
load(paste0(here::here(),file="/data-prep/data-products/cen_data_2011_hamiltononly.Rdata"))
cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
rm("cen_data_2011_1")
load(paste0(here::here(),file="/data-prep/data-products/cen_data_2016_hamiltononly.Rdata"))
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")
load(paste0(here::here(),file="/data-prep/data-products/cen_data_2021_hamiltononly.Rdata"))
cen_data_2021_hamiltononly <- cen_data_2021_1 
rm("cen_data_2021_1")

@@

#adding $ URBAN to the 2021 census DA file... retrieved from parcel points. No other year of parcel points provide this information 
load(paste0(here::here(),file="/data-prep/data-products/DA_2021_URBvsRUR.Rdata"))
cen_data_2021_hamiltononly <- cen_data_2021_hamiltononly |> left_join(DA_2021_URBvsRUR, by="GeoUID")
```

@@

```{r}
community_car_perc_2011 <- avail_2011 |>
  group_by(COMMUNITY) |> 
  summarize(TAZOrig_car_perc = mean(TAZOrig_car_perc, na.rm=T))

@@

```{r, eval=FALSE}
#check to make sure that V_i is equal to the total capacity in Hamilton. It is!
avail_2011$V_ij %>% sum()
SCHOOLS$OTGC2011 %>% sum(na.rm = TRUE)

@@

```{r include=FALSE}
#the tts origin-destination trips, by mode
load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2011.Rdata"))
load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2016.Rdata"))

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() +
  geom_histogram(data = od_trips_perc_tt, 
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey70",
                 fill = "grey70",
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8, alpha=0.6) + 
  xlab("Estimated travel time associated with TTS OD flows (minutes)") +
  ylab("Density") +
  scale_x_continuous(breaks = seq(0,45,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  theme_classic() +
  theme(axis.title = element_text(size = 11),
        panel.border = element_rect(colour = "grey80", fill=NA),
        strip.text = element_text(size=11)) 
TLD_empirical_wtheo_curve_plot
ggsave(paste0(here::here(),"/Figures/Fig5.tiff"), height=4.5, width=6, dpi=600)
```

@@

The amount of services accessible within communities oscillate: economic and demographic shifts, rural and urban configurations and political in/action all induce change. These forces spur pressures to reorganization and consolidation of public facilities that provide services, potentially improving access for some but not for all [@Christiaanse_2020; @Rosik_PulawskaObiedowska_Goliszek_2021]. Schools are a compelling case study in this respect. In many communities globally, school closure and consolidation policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019]. Research has highlighted that these polices have knock-on implications, including the weakening of communities' social infrastructure [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014] and negative household-level fiscal impacts such as a reduction of property values [@MerrallWhatsASchool2024]. Operationally, savings may be experienced but they are paid at different levels by students, families, communities and society.

A school offers many things to a community: but one dimension is the travel to and from school, often overlooked by decision-makers considering school closure and consolidation policies [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017]. Schools and residential locations often proximately co-located, the journey-to-school is an excellent opportunity for physical activity [@desjardinsFramingActive2024]. However, closures exchange the physical activity of students' realized active trips or "reserved" active trips [@morencyStepsReserve2009; @morencyHowMany2007] (e.g., trips that could be made active) for motorized travel [@rongReviewResearchLowcarbon2022]. Indeed, motorized travel to school is the largest contributor of carbon emissions from personal daily travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024]. Despite these societal implications, many communities are closing and/or consolidating schools [@sageman2022]. For instance, 65% of comprehensive schools (e.g. grades 1 through 9) in rural Finland since 1990 closed [@AuttiHyryBeihammer2014], 10% of elementary schools (kindergarten through grade 8) in Chicago, USA in 2013 [@Lee_Lubienski_2017], and 7% of primary and lower secondary schools (kindergarten through grade 9) were closed in Denmark between 2010-2011 [@Beuchert_Humlum_Nielsen_Smith_2018].

In this paper, we explore the case study of Hamilton, a mid-sized city in Canada that has pursued similar school closure and consolidation policies. We detail the Canadian political context that incentived these policies and quantify their transportation implications using data that constructs where students may have lived and the capacity of schools in 2011, 2016 and 2021. To quantify transportation implications, we use a novel multimodal extension of a singly-constrained accessibility measure known as spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024]. Spatial availability is used to estimate active (i.e., walking) and motorized (i.e., car) accessibility considering the capacity of each school, and the residential location and quantity of school-aged population, and travel times from residential locations to schools for three time periods. In this way, this work contributes a empirical findings documenting how government austerity hurts public education, along with a methodology that results in interpretable accessibility values that can be compared across modes, the City, and three time periods.

To foreshadow the paper's conclusions, our findings indicate that school seats have became less spatially available in the city. This change happened unevenly, with certain communities being disproportionately affected. The school closure and consolidation policy pursued may have achieved cost savings in facility operation and maintenance. However, it has come at a cost to families, students, and communities, resulting in lost opportunities for healthier active travel and a potential increase in motorized travel. We investigate: were the closures and consolidations worth the operational savings? The methods used in this paper can be applied by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy. Furthermore, to increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c]. Accordingly, all associated code and data is publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts).

@@

The examination of access to public services is a well established area of the transportation literature: from parks [@elmurrWalkingAccessibilityParks2021] to healthcare [@pereiraGeographicAccessCOVID192021] and associated dimensions of inequities [@tiznadoaitkenPublicTransportAccessibility2021; @bittencourtEvaluatingAccessibilityAvailability2023; @kelobonye2020measuring], spatial accessibility has been explored in cases studies worldwide.

Accessibility to schools has been a public service explored across several transportation dimensions. For instance, @marquesAccessibilityPrimarySchools2020 demonstrates that lower-income neighbourhoods, compounded by other intersecting dimensions of disadvantage, tend to have lower accessibility to schools compared to more advantaged neighbourhoods in Portugal. Similar conclusions are drawn in West Virginia, US by @Talen_2001, England by @burgess2011parental, Baton Rouge, US by @williamsDisparitiesAccessibilityPublic2014, São Paulo and Curitiba, Brazil by @morenomonroyPublicTransportSchool2018, @pizzolQualifyingAccessibilityEducation2021, and @bittencourtEvaluatingAccessibilityAvailability2023, and Santiago, Chile by @tiznadoaitkenPublicTransportAccessibility2021. Research has also demonstrated that when school closures occurs, what schools end up closing are not a politically neutral decision. For instance, these spatial trends are explored in the context of England by @pinchChangingGeographyPreschool1987, areas of declining school-aged populations in Dreseden, Germany by @mullerAssessmentSchoolClosures2011, rural Finnish communities by @AuttiHyryBeihammer2014, and Chicago, US by @Lee_Lubienski_2017.

The methods applied to measure school accessibility have been varied, from simple indicators to complex metrics. Beginning from more simple, @marquesAccessibilityPrimarySchools2020 present indicators of school seats per student along with the shortest distance from origin to schools within a school catchment, methods also used in @Talen_2001 and @burgess2011parental. Authors make refinements to these methods; @mullerAssessmentSchoolClosures2011 considers travel time by public transit from origins to schools and @pizzolQualifyingAccessibilityEducation2021 considers travel times by different modes and multiple schools from a single origin along with other indicators related to school quality. By contrast, @williamsDisparitiesAccessibilityPublic2014, @Lee_Lubienski_2017, @morenomonroyPublicTransportSchool2018, @tiznadoaitkenPublicTransportAccessibility2021, and @bittencourtEvaluatingAccessibilityAvailability2023 use more complex approaches such as the two-step floating catchment area (2SFCA). The 2SFCA is popular in measure in evaluating access to healthcare services [@paez2019] and yields a supply-to-demand ratio that accounts for the population demanding a service and the quantity that is supplied within a certain distance range [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003]. The consideration of the supply of service is pertinent in the case of schools and is acknowledged to as a limitation to using simpler accessibility measures [@pizzolQualifyingAccessibilityEducation2021].

In addition to competition for school-seats, theoretically, the advantage students using faster modes (e.g., motorized modes) offer in terms of a wider range in school-seat choice is also important. Though some of the reviewed works using 2SFCA approaches consider multiple modes, explicit attention to the modal _advantage_ has not been discussed. To this end, in this work we use spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024]. Spatial availability allows for the representation of how many 'school-seats' are spatially available to motorized- and active-trip taking elementary school-aged student population at a spatial-level, facilitating interpretable comparison across years. Further, spatial availability can be represented as school-seats per capita, mathematically equivalent to the supply-to-demand ratio of the 2SFCA (see appendix in @soukhovIntroducingSpatialAvailability2023).

@@

Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022]. The city's public English education is covered by the Hamilton-Wentworth District School Board (HWDSB). In 2013, the HWDSB released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of its elementary schools (attended by students aged approximately 5 to 14) would be subject to evaluations, including a "Pupil Accommodation Review" [@hwdsbLongTermFacilities2013]. This Master Plan and the slated Pupil Accommodation Reviews prompted public outcry, as the outcome of Pupil Accommodation Reviews are typically schools closures and consolidations [@craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].

Like all school boards in Ontario, HWDSB relies on funding from the province which is distributed through a formula based on pupil enrolment [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012]. However, it has been demonstrated that this funding formula has become insufficient to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015]. Further, existing "top-up" programs to assist in operation and maintenance costs (e.g., School Facilities Operation and Renewal Grant) were entirely phased out by 2018 [@tdsbLiftingMinistryEducations2024]. Within this funding backdrop, the Pupil Accommodation Review guidelines outlined a "community-focused" process to assessing the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006]. However in practice, critical residents felt the outcomes of the Pupil Accommodation Reviews were predetermined [@thompsonItJustGalls2024]: older schools were slated to be closed, and a wholesome breakdown of costs to repair and maintain schools were not made available [@kleinhuisSeriousConcernsSchool2013]. Other critics pointed to unbalanced incentives motivating the HWDSB: "By closing three schools, the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. The Ministry's framework promotes the closing and consolidation of schools rather than cost-effective solutions for keeping older schools open [@kleinhuisSeriousConcernsSchool2013].

The HWDSB and the Hamilton-Wentworth Catholic District School Board (HWCDSB), which together provide public English education to the majority of school-aged children in Hamilton [@faoOntarioSchoolBoards2023], were particularly impacted. While Accommodation Reviews were a province-wide trend during this period, Hamilton's school boards underwent an unprecedented number of reviews compared to other boards in other Ontario municipalities. [@seasons2014ARCCatalogues2014]. In the wave of Pupil Accommodation Reviews between 2013 and 2016, `r SCHOOLS %>% filter(Status == "Removed") %>% pull(SchoolID) %>% length()` of the `r SCHOOLS %>% filter(Year == "2011 and 2016" | Year == "2011") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton. Likely in part due to the public unpopularity of school closures, in 2017, the provincial government introduced a moratorium on beginning new Pupil Accommodation Reviews. In the following years, the HWDSB addressed its backlog and closed `r SCHOOLS %>% filter(Status_2 == "Removed") %>% pull(SchoolID) %>% length()` more schools between 2017 to 2021 [@hwdsb2023LongTermFacilities2023]. Though elementary school locations declined by `r -((SCHOOLS %>% filter(Year_2 != "2016") %>% pull(SchoolID) %>% length() - SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length())/SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length()) %>% scales::percent()` between 2011-2021, HWDSB intends to conduct future Pupil Accommodation Reviews as initially scheduled, waiting until the end of the moratorium before dates are assigned [@hwdsb2023LongTermFacilities2023].

>>>>

---
title: "10 years of school closures and consolidations: the impact on school accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
format:
  nature-pdf: 
    journal: "default"
    keep-tex: true
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca"
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing motorized travel and increasing active travel are top priorities for many communities. However, these priorities clash against many government's reluctance to invest in public services, including schools. Public school boards are increasingly asked to do more with less, a directive often accomplished by closing and consolidating schools. Alas, this policy tends to reduce the proximity of schools for many students, thus reducing the chances for active travel. In this paper we study the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced numerous elementary schools closures between 2011 and 2021. Our analysis evaluates how school closures and consolidations reshaped accessibility for elementary school students by motorized and active travel. In this task we use spatial availability, a singly-constrained multimodal measure of spatial accessibility. This method's proportional allocation mechanisms allow us to make consistent comparisons in accessibility between time periods. Analysis is conducted at the parcel level, considering on-the-ground-capacity (OTGC) of schools and observed home-to-school trip lengths for motorized and active modes. Our findings reveal that overall spatial availability in the city declined, and while the school consolidation policy pursued by the province may have achieved cost savings in facility operation and maintenance, it imposes ongoing costs on students, families, and communities through reduced opportunities for physical activity and entrenched reliance on motorized travel.

@@

```{r, include=FALSE}
load(paste0(here::here(),file="/data-prep/data-products/cen_data_2011_hamiltononly.RData"))
cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
rm("cen_data_2011_1")
load(paste0(here::here(),file="/data-prep/data-products/cen_data_2016_hamiltononly.RData"))
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")
load(paste0(here::here(),file="/data-prep/data-products/cen_data_2021_hamiltononly.RData"))
cen_data_2021_hamiltononly <- cen_data_2021_1 
rm("cen_data_2021_1")

@@

#adding $ URBAN to the 2021 census DA file... retrieved from parcel points. No other year of parcel points provide this information 
load(paste0(here::here(),file="/data-prep/data-products/DA_2021_URBvsRUR.RData"))
cen_data_2021_hamiltononly <- cen_data_2021_hamiltononly |> left_join(DA_2021_URBvsRUR, by="GeoUID")
```

@@

```{r cache=FALSE}
community_car_perc_2011 <- avail_2011 |>
  group_by(COMMUNITY) |> 
  summarize(TAZOrig_car_perc = mean(TAZOrig_car_perc, na.rm=T))

@@

```{r, include=FALSE}
#check to make sure that V_i is equal to the total capacity in Hamilton. It is!
avail_2011$V_ij %>% sum()
SCHOOLS$OTGC2011 %>% sum(na.rm = TRUE)

@@

```{r include=FALSE}
#the tts origin-destination trips, by mode
load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2011.RData"))
load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2016.RData"))

@@

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() +
  geom_histogram(data = od_trips_perc_tt, 
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey70",
                 fill = "grey70",
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", linewidth=0.8, alpha=0.6) + 
  xlab("Estimated travel time associated with TTS OD flows (minutes)") +
  ylab("Density") +
  scale_x_continuous(breaks = seq(0,45,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
  theme_classic() +
  theme(axis.title = element_text(size = 11),
        panel.border = element_rect(colour = "grey80", fill=NA),
        strip.text = element_text(size=11)) 
TLD_empirical_wtheo_curve_plot
ggsave(paste0(here::here(),"/Figures/Fig5.tiff"), height=4.5, width=6, dpi=600)
```

@@

Public services accessible within communities tend to oscillate: economic and demographic shifts, rural and urban configurations and political in/action all induce change. These forces spur pressures to reorganize and consolidate public facilities, potentially improving access for some but not for all residents in a community [@Christiaanse_2020; @Rosik_PulawskaObiedowska_Goliszek_2021]. Public schools are a compelling case study in this respect. Across the globe, school closure and consolidation policies have been pursued by governments to reduce the cost of running public education systems  [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019]. Research has highlighted that these polices have knock-on implications, including the weakening of communities' social infrastructure [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014] and negative household-level fiscal impacts such as a reduction of property values [@MerrallWhatsASchool2024]. Operationally, savings may be experienced by governments, but they are paid at different levels by students, families, communities, and ultimately society in general.

A school offers many things to a community, and one relevant dimension is how it shapes travel. This dimension is often overlooked by decision-makers tasked with implementing school closure and consolidation policies [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017]. When schools and residential locations are co-located, the journey-to-school presents an excellent opportunity for physical activity [@desjardinsFramingActive2024]. However, school closures exchange the opportunities for active trips by students [@morencyStepsReserve2009; @morencyHowMany2007] for the near certainty of motorized travel [@rongReviewResearchLowcarbon2022]. Indeed, motorized travel to school is the largest contributor of carbon emissions from personal daily travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024]. Alas, these impacts are not often considered in policy when schools close and/or consolidate [@sageman2022], a relatively common occurrence globally. For instance, 65% of comprehensive schools (e.g. grades 1 through 9) in rural Finland since 1990 closed [@AuttiHyryBeihammer2014], 10% of elementary schools (kindergarten through grade 8) in Chicago, USA in 2013 [@Lee_Lubienski_2017], and 7% of primary and lower secondary schools (kindergarten through grade 9) were closed in Denmark between 2010-2011 [@Beuchert_Humlum_Nielsen_Smith_2018].

In this paper, we investigate the case of Hamilton, a mid-sized city in Canada that has been forced by the province of Ontario to pursue school closure and consolidation policies. We <!--detail the Canadian political context that incentivized these policies and--> quantify the transportation implications of these policies using the capacity of schools in 2011, 2016 and 2021, after several waves of school closures and consolidations. To this end, we use a novel multimodal extension of a singly-constrained accessibility measure known as spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024]. We use this measure to estimate active (i.e., walking) and motorized (i.e., car) accessibility while constraining the supply of the service to the capacity schools. In this way, this work furnishes evidence to document how government austerity hurts public goals (such as physical activity) and locks school districts and families into motorized patterns of travel for the foreseeable future.

To foreshadow the paper's conclusions, our findings indicate that school seats have became less available in the city. This change happened unevenly, with certain communities being disproportionately affected. The school closure and consolidation policy pursued by the government may have achieved cost savings in facility operation and maintenance. However, it has come at a cost to families, students, and communities, resulting in lost opportunities for healthier life-styles and an inevitable increase in motorized travel [@hammerWhy2012; @roseKids2013; @buckWhy2019]. To increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c]. For this reason, all associated code and data are publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts).

@@

The examination of accessibility to public services is well-established, including studies of parks [@reyesalking2014; @elmurrWalkingAccessibilityParks2021], healthcare [@pereiraGeographicAccessCOVID192021], and public transportation [@moniruzzamanAccessibility2012; @tiznadoaitkenPublicTransportAccessibility2021], among many other services [e.g., @bittencourtEvaluatingAccessibilityAvailability2023; @kelobonye2020measuring].

Schools are a service that has been studied across several transportation dimensions. For instance, @marquesAccessibilityPrimarySchools2020 demonstrate that lower-income neighbourhoods in Portugal tend to have lower accessibility to schools compared to more advantaged neighbourhoods. Similar conclusions are drawn in West Virginia, US by @Talen_2001, England by @burgess2011parental, Baton Rouge, US by @williamsDisparitiesAccessibilityPublic2014, São Paulo and Curitiba, Brazil by @morenomonroyPublicTransportSchool2018, @pizzolQualifyingAccessibilityEducation2021, and @bittencourtEvaluatingAccessibilityAvailability2023, and Santiago, Chile by @tiznadoaitkenPublicTransportAccessibility2021. Research has also demonstrated that school closures are seldom politically neutral decisions. For instance, spatial trends were explored in the context of England by @pinchChangingGeographyPreschool1987, areas of declining school-aged populations in Dreseden, Germany by @mullerAssessmentSchoolClosures2011, rural Finnish communities by @AuttiHyryBeihammer2014, and Chicago, US by @Lee_Lubienski_2017.

The methods applied to measure school accessibility have also been varied. For example, @marquesAccessibilityPrimarySchools2020 present indicators of school seats per student along with the shortest distance from origin to schools within a school catchment. Similar methods are also used in @Talen_2001 and @burgess2011parental. Several authors propose refinements to these methods; @mullerAssessmentSchoolClosures2011, for instance, considers travel time by public transit from origins to schools, whereas @pizzolQualifyingAccessibilityEducation2021 considers travel times by different modes and multiple schools from a single origin along with other indicators related to school quality. By contrast, @williamsDisparitiesAccessibilityPublic2014, @Lee_Lubienski_2017, @morenomonroyPublicTransportSchool2018, @tiznadoaitkenPublicTransportAccessibility2021, and @bittencourtEvaluatingAccessibilityAvailability2023 use more complex approaches such as the two-step floating catchment area (2SFCA). The 2SFCA is popular in measure in evaluating access to healthcare services [@paez2019] and yields a supply-to-demand ratio that accounts for the population demanding a service and the quantity that is supplied within a certain distance range [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003]. The consideration of the supply of service is pertinent in the case of schools and is acknowledged to as a limitation to using simpler accessibility measures [@pizzolQualifyingAccessibilityEducation2021].

In addition to competition for school-seats, theoretically, the use of faster modes (e.g., motorized modes) offers an advantage in terms of a wider range in school-seat choice. Though some of the works reviewed consider multiple modes, explicit attention to the modal _advantage_ has not been discussed. To this end, in this work we use spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024]. Spatial availability allows for the representation of the number of 'school-seats' that are spatially available to students by mode of transportation (i.e., motorized and active travel), considering the school-aged student population by zone. Further, spatial availability can be represented as school-seats per capita, mathematically equivalent to the supply-to-demand ratio of the 2SFCA (see appendix in @soukhovIntroducingSpatialAvailability2023), which increases the interpretability of the analysis, and the comparability between years.

@@

Hamilton is a mid-size city (\~570,000 pop) in the province of Ontario, Canada, with rural, suburban and urban characteristics [@governmentofcanadaProfileTableCensus2022]. The city's public English education system is managed by the Hamilton-Wentworth District School Board (HWDSB) and the Hamilton-Wentworth Catholic District School Board (HWCDSB). These two boards, like many in the province have faced fiscal pressures from the province. For example, in 2013, the HWDSB released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013], which nominally meant to ensure "equitable, affordable and sustainable learning facilities". This plan indicated that 80% of its elementary schools (attended by students aged approximately 5 to 14) would be subject to evaluations, including a "Pupil Accommodation Review" [@hwdsbLongTermFacilities2013]. The announcement promptly led to a public outcry, since the outcome of Pupil Accommodation Reviews historically translated into school closures and consolidations [@craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].

School boards in Ontario rely on funding from the province which is distributed through a formula based on pupil enrollment [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012]. However, this funding formula has demonstrably become insufficient to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015]. Further, existing "top-up" programs to assist in operation and maintenance costs (e.g., School Facilities Operation and Renewal Grant) were entirely phased out by 2018 [@tdsbLiftingMinistryEducations2024]. Against this backdrop, the Pupil Accommodation Review guidelines outlined a "community-focused" process for assessing the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006]. However in practice, critical residents felt the outcomes of the Pupil Accommodation Reviews were predetermined [@thompsonItJustGalls2024]: older schools were slated to be closed, and a wholesome breakdown of costs to repair and maintain schools was not made available [@kleinhuisSeriousConcernsSchool2013]. The process introduced unbalanced incentives for school boards: as a concerned resident pointed at the time, "[b]y closing three schools, the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. The Ministry's framework promotes the closing and consolidation of schools rather than cost-effective solutions for keeping older schools open [@kleinhuisSeriousConcernsSchool2013].

Hamilton's two school boards (HWDSB and HWCDSB), which together provide public English education to the majority of school-aged children in Hamilton [@faoOntarioSchoolBoards2023], were particularly impacted by fiscal pressures. While Accommodation Reviews were a province-wide trend during this period, Hamilton's school boards underwent an unprecedented number of reviews compared to boards in other Ontario municipalities. [@seasons2014ARCCatalogues2014]. In the wave of Pupil Accommodation Reviews between 2013 and 2016, `r SCHOOLS %>% filter(Status == "Removed") %>% pull(SchoolID) %>% length()` of the `r SCHOOLS %>% filter(Year == "2011 and 2016" | Year == "2011") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton. Likely in part due to the unpopularity of school closures, in 2017, the provincial government introduced a moratorium, which paused any scheduled Pupil Accommodation Reviews. This created a backlog of Accommodation Reviews that HWDSB caught up to in subsequent years, when it closed `r SCHOOLS %>% filter(Status_2 == "Removed") %>% pull(SchoolID) %>% length()` more schools between 2017 to 2021 [@hwdsb2023LongTermFacilities2023]. Though elementary school locations declined by `r -((SCHOOLS %>% filter(Year_2 != "2016") %>% pull(SchoolID) %>% length() - SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length())/SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length()) %>% scales::percent()` between 2011-2021, HWDSB intends to conduct future Pupil Accommodation Reviews as initially scheduled, waiting until the end of the moratorium before dates are assigned [@hwdsb2023LongTermFacilities2023].
